---
title: "HW2 BIOST540"
author: "Meixin Zhang"
date: "4/21/2021"
output: word_document
---
## Part A
```{r}
rm(list=ls())
library(reshape2)
library(ggplot2)
library(dplyr)
library(nlme)
cholesterol <- read.csv("~/Documents/MS-2nd-year/Spring/BIOST540/HW/cholesterol.csv")
```

## Question 1

```{r}
cholesterol<-cholesterol[,c("group","id","y1","y2","y3","y4","y5")]
choles_long <- melt(cholesterol, id=c("id","group"))
names(choles_long)<-c("id","group","time","value")
choles_long$time <- as.numeric(gsub("y","",choles_long$time))
choles_long$group[choles_long$group==1]<-"Treatment"
choles_long$group[choles_long$group==2]<-"Control"
choles_means <- aggregate(value ~ time + group, data = choles_long, FUN=mean)
p <- ggplot(data = choles_means, aes(x = time, y = value, group = group, 
                                    col = group))
p + geom_line() + scale_color_manual(values=c('blue','red')) + 
  geom_point() + theme_bw() + ylab("Serum cholesterol (mg/dL)") + xlab("Time points")
```

## summary statistics
```{r}
by(cholesterol[,-c(1,2)], INDICES = cholesterol$group, FUN=summary)
```
## Estimated correlation matrix after subtracting out means by group and month
```{r}
choles_mean <- cholesterol %>% group_by(group) %>% 
  summarise(y0mean = mean(y1), y6mean=mean(y2), 
            y12mean=mean(y3,na.rm = TRUE), y20mean=mean(y4,na.rm = TRUE),
            y24mean=mean(y5,na.rm = TRUE)) 
cholesterol <- merge(cholesterol, choles_mean, by="group")
cholesterol$y0.resid <- cholesterol$y1 - cholesterol$y0mean
cholesterol$y6.resid <- cholesterol$y2 - cholesterol$y6mean
cholesterol$y12.resid <- cholesterol$y3 - cholesterol$y12mean
cholesterol$y20.resid <- cholesterol$y4 - cholesterol$y20mean
cholesterol$y24.resid <- cholesterol$y5 - cholesterol$y24mean
choles_complete<-cholesterol[rowSums(is.na(cholesterol))==0,]
cor(choles_complete[,c("y0.resid", "y6.resid", "y12.resid", "y20.resid","y24.resid")])
```

## Estimated correlation matrix by treatment groups
```{r}
by(choles_complete[,c("y0.resid", "y6.resid", "y12.resid", "y20.resid", "y24.resid")], 
   INDICES = choles_complete$group, FUN=cor)
```

## Question 3
```{r}
m.reml<-gls(value~group*as.factor(time),
            data=choles_long[rowSums(is.na(choles_long))==0,],
            method="REML",
            correlation=corSymm(form = ~time | id),
            weights=varIdent(form= ~1 | time))
summary(m.reml)
anova(m.reml)
```

## Question 4
```{r}
choles_long$month[choles_long$time==1]<-0
choles_long$month[choles_long$time==2]<-6
choles_long$month[choles_long$time==3]<-12
choles_long$month[choles_long$time==4]<-20
choles_long$month[choles_long$time==5]<-24
m.reml2<-gls(value~group*month,
             data=choles_long[rowSums(is.na(choles_long))==0,],
             method="REML",
             correlation=corSymm(form = ~time | id),
             weights=varIdent(form= ~1 | month))
summary(m.reml2)
```

## Part B
```{r}
data(Orthodont)
Orthodont$id<-cumsum(!duplicated(Orthodont$Subject))
```

## Question 2
## OLS
```{r}
lm<-lm(distance ~ as.factor(age)*Sex, 
            data = Orthodont)
summary(lm)
```

## GLS, unstructured/symmetric correlation matrix, heteroscedasticity, REML 
```{r}
Orthodont$time<-ifelse(Orthodont$age==8,1,
                       ifelse(Orthodont$age==10,2,
                              ifelse(Orthodont$age==12,3,4)))
reml_2<-gls(distance ~ Sex*as.factor(age),
            data = Orthodont,
            method = "REML",
            correlation = corSymm(form = ~time | Subject),
            weights = varIdent(form = ~1 | as.factor(age)))
summary(reml_2)
```

## GLS, exchangeable/compound symmetric correlation matrix, homoscedasticity, REML 
```{r}
reml_3<-gls(distance ~ Sex*as.factor(age), 
              data=Orthodont, 
              method="REML",
              correlation=corCompSymm(form = ~time | id))
summary(reml_3)
```

## LMM, random intercepts, REML 
```{r}
lmm1 <- lme( distance ~ Sex*as.factor(age), 
             method = "REML", data = Orthodont, 
             random = reStruct( ~ 1 | id, pdClass="pdDiag", REML=F))
summary(lmm1)
```

## LMM, random intercepts + slopes (correlated), REML
```{r}
lmm2 <- lme( distance ~ Sex*as.factor(age), 
             method = "REML", data = Orthodont, 
             random = reStruct( ~ 1 + age | id, pdClass="pdSymm", REML=F))
summary(lmm2)
```

